Variational Monte Carlo on bosonic systems

### 2018-04-03 project_1/article/article.tex
Start on theory of MC integration
One-body results

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due time). This
        alteration of the algorithm is dependent on the \emph{drift force} of
        the system\footnote{Also known as the \emph{quantum force}}.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$ as the Monte Carlo sampling involves
        choosing a single particle at a time.

\subsection{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
    hard-shell interaction between the particles and that we use a spherical
    potential trap.  We thus get the trial  wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\para{
            -\alpha |\vf{r}_i|^2
        },
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplacian we can compute the kinetic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, $\hbar = c = 1$, and with the mass set to unity, $m = 1$,
    this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \omega/2$\footnote{Note that $\alpha$
    is required to be positive.} we get a stable value for the local energy.
    \begin{align}
        E_L(\vf{r})
        &\equiv
        E_L
        =
        \frac{\omega d N}{2},
        \label{eq:minimum_energy_simple}
    \end{align}
    which is the stable minimum for $\alpha = \omega/2$.

    \subsubsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
            \\
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
            \label{eq:exact_energy}
        \end{align}
        By minimizing this expression with respect to $\alpha$ yields
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}
        which in natural units reduces to $\alpha_0 = \omega/2$. The energy at
        this value of $\alpha$ (in natural units) is then
        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2},
        \end{align}
        in perfect accordance with \autoref{eq:minimum_energy_simple} which
        yields high hopes for the latter equation and the expected minimum.

@@

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}.

>>>>

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will use the Metropolis-Hastings algorithm with the additions of
        \emph{Importance Sampling}  (which will be discussed in due time). This
        alteration of the algorithm is dependent on the \emph{drift force} of
        the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

\subsection{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$, i.e. there is no hard-shell
    interaction between particles and the particles are confined in a spherical trap.
    We thus get the trial  wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplacian we can compute the kinetic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the sum over all particles
    disappears.

    \subsubsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}.
        \end{align}

        % TODO: Add the computation of this expression
        The final result for the energy is
        \begin{align}
            E(\alpha)
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
            \label{eq:exact_energy}
        \end{align}
        By minimizing this expression, i.e., finding the derivative of the
        energy with respect to $\alpha$ and equating this to zero, yields the
        expected minimum of variational energy to be
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}
        which in natural units reduces to $\alpha_0 = \half\omega$. The energy at
        this value of $\alpha$ (in natural units) is then
        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2}.
        \end{align}

@@

            For this optimal, interacting $\alpha$ we have computed the expected
            energy and standard deviation for systems of $n = \{10, 50, 100\}$
            particles. These results are shown in \autoref{tab:10_optimal}, the 
            non-interacting case, excluding Jastrow factor is included as a 
            comparison.

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}. The results for a system consisting
            of $n=100$ particles are displayed in \autoref{fig:one_body_density_100}.
            In this figure we see the densities for for a system confined to a 
            perturbed elliptic trap with and without interaction. We see that for the
            interacting case, the one-body density is skewed towards a larger 
            $\abs{\vf{r}}$, as is expected. The `ìdeal'' density, a system with spherical
            trap, is also included in the figure.
